Dependent Object Types

### 2012-08-05 article.tex
revision: one pass through

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types calculus (DOT). DOT models Scala's
path-dependent types and abstract type members, as well as its mixture
of nominal and structural typing through the use of reﬁnement
types. It makes no attempt to model inheritance or mixing
composition. The calculus does not model what's currently in Scala: it
is more normative than descriptive.

@@

\category{D.3.3}{Language Constructs and Features}{Abstract data types, Classes and objects, polymorphism}
\category{D.3.1}{Formal Deﬁnitions and Theory}{Syntax, Semantics}
\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{}
\category{F.3.3}{Studies of Program Constructs}{Object-oriented constructs, type structure}
\category{F.3.2}{Semantics or Programming Languages}{Operational semantics}

@@

The types in DOT are denoted by letters $S$, $T$, $U$, or $V$. They consist of the following:
\begin{itemize}
\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
         other declarations in the same type by selecting from $z$.
\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
\item[-] A top type $\Top$, which corresponds to an empty object.
\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
\end{itemize}
A subset of types $T_c$ are called {\em concrete types}. These are type selections
$p.L_c$ of class labels,
the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.

There are only three forms of declarations in DOT, which are both part
of refinement types.  A value declaration $\ldecl l T$ introduces a
field with type $T$.  A method declaration $\mdecl m S U$ introduces a
method with parameter of type $S$ and result of type $U$. A type
declaration $\Ldecl L S U$ introduces a type member $L$ with a lower
bound type $S$ and an upper bound type $U$. There are no type aliases,
but a type alias can be simulated by a type declaration $\Ldecl L T T$
where the lower bound and the upper bound are the same type $T$.

@@

Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
terms $t$ and stores $s$, where stores map variables to constructors.
There are three main reduction rules: Rule (\textsc{msel}) rewrites a
method invocation $y {m_i} v$ by retrieving the corresponding method
definition from the store, and performing a substitution of the
argument for the parameter in the body. Rule (\textsc{sel}) rewrites a
field selection $x.l$ by retrieving the corresponding value from the
store. Rule (\textsc{new}) rewrites an object creation $\new x c t$ by
placing the binding of variable $x$ to constructor $c$ in the store
and continuing with term $t$.  These reduction rules can be applied
anywhere in a term where the hole $[\,]$ of an evaluation context $e$
can be situated.

@@

Rules (\textsc{sel}) and (\textsc{msel}) type field selections and
method invocations by means of an auxiliary membership relation $\ni$,
which determines whether a given term contains a given declaration as
one of its members. The membership relation is defined in
\figref{dot-two} and is further explained below.

@@

\subsection{Membership}

\figref{dot-two} presents typing rules for membership and
expansion.  The membership judgement $\Gamma \ts t \ni D$ states that
in environment $\Gamma$ a term $t$ has a declaration $D$ as a member.
There are different rules for paths and general terms.  Rule
(\textsc{path-$\ni$}) applies to paths $p$ that have a refinement type
$T \refine z {\seq D}$ as their type. Members of $p$ are then all
definitions $D$ in the refinement, where any use of the self reference
$z$ is replaced by the path $p$ itself.  Rule (\textsc{and-$\ni$})
allows to merge two member definitions (of the same label) by
conjoining them with $\tand$. Rule (\textsc{term-$\ni$}) establishes
the members of general terms $t$ of type $T$ by introducing a dummy
variable $z$ of type $T$ and then using the rules for path membership
on $z$. It must hold that $z$ itself does not form part of the
resulting member $D$.

@@

The expansion relation $\expand$ is needed to typecheck the complete
set of declarations carried by a concrete type that is used in a
\textbf{new}-expression. Expansion is also used in subtyping
refinements on the right (see \figref{dot-sub}).

@@

\figref{dot-sub} defines the subtyping judgement $\Gamma \ts S
\sub T$ which states that in environment $\Gamma$ type $S$ is a
subtype of type $T$. Subtyping is regular with respect to $\wfe$: if
type $S$ is a subtype of type $T$, then $S$ and $T$ are well-formed
and expanding. Though this regularity limits our calculus to
$\wfe$-types, this limitation allows us to show that subtyping is
transitive, as discussed in section~\ref{dot-preservation}.

@@

We prove that the DOT calculus is type-safe using logical relations in
section~\ref{dot-type-safety}. However, we first tried to prove the
calculus type-safe using the standard theorems of preservation and
progress~\cite{soundness,tapl}. Unfortunately, for the calculus as
presented, and any of its variants that we devised, preservation
doesn't hold. In this section, we review some of the most salient
counterexamples to preservation that we found. These counterexamples
have been checked with PLT Redex.

@@

Now, consider the term
\begin{align*}
&\mlnew u {X \ldefs{ l = u }}{
\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {\app {(\abs x X x)} u})}.l
}
\end{align*}

The term type-checks because the term $t=\app {(\abs y {\Top \tfun Y}
  {\app y u})} {(\abs d \Top {\app {(\abs x X x)} u)}}$ has type
$Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
$t$ eventually steps to $\app {(\abs x X x)} u$ which has type $X$,
so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
self-reference ($z.L_a$).

@@

@@

Originally, subtyping was not regular with respect to $\wfe$. Roughly,
all the $\wfe$ preconditions in subtyping were dropped. In this
broader calculus, subtyping transivitiy doesn't hold, because of the
rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.

@@

The idea is to start with a function from $S \tfun S$ and cast it
successively to $S \tfun T$ then $S \tfun U$. To typecheck the
expression initially, we need to check $S \sub T$ and $T \sub
U$. After some reduction steps, the first few casts vanish, and the
reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
we need to check $S \sub U$. Thus, we need subtyping transitivity: $S
\sub T$ and $T \sub U$ implies $S \sub U$.

@@

Now, $f$ is a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
  x)}$ is stuck (and, rightfully, doesn't typecheck). But we can use
narrowing to create a counterexample to type safety: $\app {(\abs g
  {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.

@@

An alternative design would have been to change the expansion of the
function type to have a special marker value label.

@@

Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
example doesn't typecheck initially because the path-equality
provisions only applies when objects are in the store, so the
application is not well-typed. But if we start preservation in a store
which has $a$, $b$, $c$ and $d$ then the application type-checks,
because, through one of the path-equality provision, we can find that
the type of $d$ is a subtype of $\Bot$. Now, of course, when we get to
$d.\mathit{foo}$, reduction fails.

\section{Conclusion}\label{conclusion}

>>>>

\begin{abstract}
We propose a new type-theoretic foundation of Scala and languages like
it: the Dependent Object Types calculus (DOT). DOT models Scala's
path-dependent types and abstract type members, as well as its mixture
of nominal and structural typing through the use of refinement
types. It makes no attempt to model inheritance or mixing
composition. The calculus does not model what's currently in Scala: it
is more normative than descriptive.

@@

\category{D.3.3}{Language Constructs and Features}{Abstract data types, Classes and objects, polymorphism}
\category{D.3.1}{Formal Definitions and Theory}{Syntax, Semantics}
\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{}
\category{F.3.3}{Studies of Program Constructs}{Object-oriented constructs, type structure}
\category{F.3.2}{Semantics or Programming Languages}{Operational semantics}

@@

The types in DOT are denoted by letters $S$, $T$, $U$, $V$, or $W$. They consist of the following:
\begin{itemize}
\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
         other declarations in the same type by selecting from $z$.
\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
\item[-] A top type $\Top$, which corresponds to an empty object.
\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
\end{itemize}
A subset of types $T_c$ are called {\em concrete types}. These are type selections
$p.L_c$ of class labels,
the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.

There are only three forms of declarations in DOT, which are all part
of refinement types.  A value declaration $\ldecl l T$ introduces a
field with type $T$.  A method declaration $\mdecl m S U$ introduces a
method with parameter of type $S$ and result of type $U$. A type
declaration $\Ldecl L S U$ introduces a type member $L$ with a lower
bound type $S$ and an upper bound type $U$. There are no type aliases,
but a type alias can be simulated by a type declaration $\Ldecl L T T$
where the lower bound and the upper bound are the same type $T$.

@@

Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
terms $t$ and stores $s$, where stores map variables to constructors.
There are three main reduction rules: Rule (\textsc{msel}) rewrites a
method invocation $\mapp y {m_i} v$ by retrieving the corresponding method
definition from the store, and performing a substitution of the
argument for the parameter in the body. Rule (\textsc{sel}) rewrites a
field selection $x.l$ by retrieving the corresponding value from the
store. Rule (\textsc{new}) rewrites an object creation $\new x c t$ by
placing the binding of variable $x$ to constructor $c$ in the store
and continuing with term $t$.  These reduction rules can be applied
anywhere in a term where the hole $[\,]$ of an evaluation context $e$
can be situated.

@@

Rules (\textsc{sel}) and (\textsc{msel}) type field selections and
method invocations by means of an auxiliary membership relation $\ni$,
which determines whether a given term contains a given declaration as
one of its members. The membership relation is defined in
\figref{dot-two} and is further explained in section~\ref{mem}.

@@

\subsection{Membership}\label{mem}

\figref{dot-two} presents typing rules for membership and expansion.
The membership judgement $\Gamma \ts t \ni D$ states that in
environment $\Gamma$ a term $t$ has a declaration $D$ as a member. The
membership rules rely on expansion. There are two rules, one for paths
(\textsc{path-$\ni$}) and one for general terms
(\textsc{term-$\ni$}). For general terms, the ``self''-reference of
the type must not occur in the resulting declaration $D$, since, to
guarantee syntactic validity, we can only substitute a path for the
``self''-reference.

@@

The expansion relation $\expand$ is needed to typecheck the complete
set of declarations carried by a concrete type that is used in a
\textbf{new}-expression. Expansion is also used by the membership
rules and in subtyping refinements on the right (see
\figref{dot-sub}).

@@

\figref{dot-sub} defines the subtyping judgement $\Gamma \ts S
\sub T$ which states that in environment $\Gamma$ type $S$ is a
subtype of type $T$. Subtyping is regular wrt $\wfe$: if
type $S$ is a subtype of type $T$, then $S$ and $T$ are well-formed
and expanding. Though this regularity limits our calculus to
$\wfe$-types, this limitation allows us to show that subtyping is
transitive, as discussed in section~\ref{dot-preservation}.

@@

We prove that the DOT calculus is type-safe using logical relations in
section~\ref{dot-type-safety}. However, we first tried to prove the
calculus type-safe using the standard theorems of preservation and
progress~\cite{soundness,tapl}. Unfortunately, for the calculus as
presented, and any of its variants that we devised, preservation
doesn't hold. In this section, we review some of the most salient
counterexamples to preservation that we found. These counterexamples
have been checked with PLT Redex~\cite{plt_redex}.

@@

Now, consider the term
\begin{align*}
&\mlnew u {X \ldefs{ l = u }}{
\app {(\abs y {\Top \tfun Y} {\app y u})} {(\abs d \Top {\cast X u})}.l
}
\end{align*}

The term type-checks because the term $t=\app {(\abs y {\Top \tfun Y}
  {\app y u})} {(\abs d \Top {\cast X u})}$ has type
$Y$, so we can apply $\textsc{term-$\ni$}$ for $l$. However, the term
$t$ eventually steps to ${\cast X u}$ which has type $X$,
so we cannot apply $\textsc{term-$\ni$}$ for $l$ because of the
self-reference ($z.L_a$).

@@

% TODO: section

@@

Originally, subtyping was not regular wrt $\wfe$. Roughly,
all the $\wfe$ preconditions in subtyping were dropped. In this
broader calculus, subtyping transivitiy doesn't hold, because of the
rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.

@@

The idea is to start with a function from $S \tfun S$ and cast it
successively to $S \tfun T$ then $S \tfun U$. To typecheck the
expression initially, we need to check $S \sub T$ and $T \sub
U$. After some reduction steps, the first few casts vanish, and the
reduced expression casts directly from $S \tfun S$ to $S \tfun U$, so
we need to check $S \sub U$.

@@

Now, $f$ was a subtype of $\Top \tfun \Top$, but $\app f {(\abs x \Top
  x)}$ was stuck (and, rightfully, didn't typecheck). But we could use
narrowing to create a counterexample to type safety: $\app {(\abs g
  {\Top \tfun \Top} {\app g (\abs x \Top x)})} f$.

@@

An alternative design would have been to change the expansion of the
function type to have a declaration for a special marker value label.

@@

Notice that $d$ has type $\Bot$ if the cast on $b$ is ignored. This
example didn't typecheck initially because the path-equality
provisions only applied when objects are in the store, so the
application was not well-typed. But if we started preservation in a store
which has $a$, $b$, $c$ and $d$ then the application type-checked,
because, through one of the path-equality provision, we could find that
the type of $d$ was a subtype of $\Bot$. Now, of course, when we got to
$d.\mathit{foo}$, reduction failed.

% TODO: more discussion subsections?

\section{Conclusion}\label{conclusion}
% TODO: section
